# The Basics of R

R is built around a few basic pieces - once you understand them,
it's easier to understand more complex commands, since everything
is built from the same basic foundations.

In programming terms, we can refer to the basic pieces that make
up R as **data types**.

## Basic data types

### Numbers

The **numeric** data type allows you to work with numbers. R can
do all the basic operations you'd expect: addition, subtraction,
multiplication and division.

At the most basic level, you can use R as a calculator by doing
standard operations like `+`, `-`, `/` (division), `*` (multiplication),
`^` (power) on numeric data:

```{r numeric_calculator, prompt=TRUE}
1 + 1
2.5 * 3
8^2
```

R also has an **integer** (whole number) data type. Integers (usually)
work exactly the same as **numeric** data, so you don't need to worry
too much about the difference for now. Integers will automatically be
converted to the more general numeric format when needed:

```{r integer_example}
# You can specify that data should be integers using "L"
1L + 1L
# Automatically converts the result to numeric
3L + 0.1
5L / 2
```

### Characters (text)

The **character** data type allows you to store and manipulate
text. Character data is created by wrapping text in either single `'` or
double `"` quotes. In programming terms, we also refer to each chunk of text
as a **string**:

```{r character_examples}
"apple"
# Note: this is still just one string. All the text, including
#   the spaces, is contained in the same chunk of text
toupper("three bananas")
# Get part of a string.
substr("carrot", 1, 3)
# Stick multiple strings together with paste0
paste0("for", "got")
```

### Logical (True/False)

The **logical** data type is used to represent the True/False result
of a logical test or comparison. These are represented by the
special values of `TRUE` and `FALSE` (basically 1 and 0, with special labels
attached to them). To do logical comparisons, you can use syntax like:

* `==`: equals. Note that you need a double equal sign to compare values,
  a single equal sign does something different.
  
```{r eq_example, prompt=TRUE}
"a" == "b"
```

* `<`, `>`: less than, greater than

```{r lt_gt_example}
3 < 4
```

* `<=`, `>=`: less than or equal to, greater than or equal to

```{r leq_geq_example}
10 >= 10
```

* `!=`: not equal to

```{r neq_example}
"hello" != "world"
```

* `!`: not, which reverses the result of another logical test:

```{r not_example, prompt=TRUE}
! (5 > 3)
```

#### Combining logicals: AND and OR

More complex logical tests can be conducted by combining multiple tests
with the **and** `&` and **or** `|` operators. 

`&` takes two logicals, e.g. `a & b`,  and returns `TRUE` if both `a` *and* `b` 
are `TRUE`, and `FALSE` otherwise.

```{r and_examples, options}
# Both conditions are true: TRUE & TRUE is TRUE
(2 > 1) & ("a" == "a")
# Only one condition is true: TRUE & FALSE is FALSE
(3 >= 3) & ("b" == "a")
```

`a | b` returns `TRUE` if either `a` *or* `b` is `TRUE`

```{r or_examples, options}
# FALSE | TRUE is TRUE
(1 > 2) | ("a" == "a")
# FALSE | FALSE is FALSE
(1 > 2) | ("a" == "b")
```

It's best to wrap each individual test in parentheses `()` to make the logic clear.

## Converting between types

Occasionally your data will be read in from a file as the wrong type.
You might be able to fix this by changing the way you read in the file,
but otherwise you should **convert** the data to the type that makes
the most sense (you might have to clean up some invalid values first).

Functions like `as.character()`, `as.numeric()` and `as.logical()` will
convert data to the relevant type. Probably the most common type conversion
you'll have to do is when `numeric` data gets treated as text and is stored
as `character`. Numeric operations like addition won't work until you fix
this:

```{r numeric_wrong_type, error=TRUE}
"1" + 1
one_fixed = as.numeric("1")
one_fixed + 1
```

## Variables: Storing Results

The results of calculations in R can be stored in **variables**: you
give a name to the results, and then when you want to look at, use
or change those results later, you access them using the same name.

You **assign** a value to a variable using either `=` or `<-` (these
are mostly equivalent, don't worry too much about the difference), putting
the variable name on the left hand side and the value on the right.

NOTE THAT EVERYTHING ON THE RIGHT HAND SIDE IS BEING ASSIGNED TO THE NAME YOU GIVE ON THE LEFT

```{r variables_example}
scale_total = 3 + 8 + 5 + 2 + 4
# Accessing saved results
scale_total
# Using saved results in another calculation
severe_disorder = scale_total >= 15
severe_disorder
# Changing a variable: this will overwrite the old value with the
#   new one, the old value won't be available unless you've
#   stored it somewhere else
scale_total = scale_total + 2
scale_total
```

**When you assign a variable, you're asking R to remember some data so you can 
use it later.** Understanding that simple
principle will take you a long way in R programming.


Variable names in R should start with a letter (`a-zA-Z`), and
can contain letters, numbers, underscores `_` and periods `.`, so
`model3`, `get.scores`, `ANX_total` are all valid variable names.

### Missing values

Functions like `sum()` and `mean()` will produce a missing
result by default if *any* values in the input are missing. Use the
`na.rm = TRUE` option (short for "`NA` remove") to ignore the missing values 
and just use the values that are available:

```{r sum_missing_example}
mean(c(1, 3, NA, 7, 9))
mean(c(1, 3, NA, 7, 9), na.rm = TRUE)
```

Other functions in R will automatically remove missing values, but
will usually warn you when they do. It's always good to check
how missing values are being treated, whatever tool you're using.

