# Dataframes and More

Once you understand the basics of R's data types, some of the more
advanced features of R start to make sense. Below, we'll cover
some of these more advanced features. In particular, we'll discuss
**data frames**, which are used to store and analyse multiple
rows and columns of data bundled together in a table.

## Factors (categorical data)

**Factors** are how R represents categorical data. They
have a fixed number of **levels**, that are set up when you first
create a factor vector:

```{r factor_example}
severity = sample(c("Moderate", "Severe"), 10, replace=TRUE)
# Setting 'levels' also sets the order of the levels
sev_factor = factor(severity, levels = c("Moderate", "Severe"))
sev_factor
```

When you're testing a factor, you use the **label** to test it:

```{r test_factor_example, options}
sev_factor == "Moderate"
```

Factors can only contain data that matches their levels,
and will produce a warning if you try to add something else:

```{r factor_non_level_example}
# Not one of the levels that was set up when the factor was
#   created:
sev_factor[1] = "Mild"
sev_factor
```

[^factor-warning]: Some of the most common problems in R result
from text data that should just be in `character` format being
stored as `factor`. `factor` should only be used if you
have categorical variables with a fixed number of levels (usually
a small number). If you have text columns, check how they've
been stored.

That said, we will convert our categorical variables in survey_one to factor
```{r}
survey_one$sex<-as.factor(survey_one$sex)
survey_one$volunteer<-as.factor(survey_one$volunteer)
```



## Dataframes!

The most common format for working with data is in a table, with
data arranged in rows and columns. R's main format for tables
of data is the **dataframe**. In a dataframe:


Most of the time, you'll read your data from a file (a spreadsheet, an
SPSS file, etc.) and it will be read in as a dataframe. 
For example, survey_one is a data frame

```{r}
class(survey_one)
```


### Accessing parts of dataframes

#### Accessing a single column {-}

To access a single column from a dataframe, you can use `$`, which will
return a single vector:

```{r df_single_column}
survey_one$sex
```

