# Better plots with `ggplot2`

R's plotting tools are very good, and in particular, there's
a great library called `ggplot2` that allows you to put together
a huge variety out of plots using a simple system and some
common pieces.

`ggplot2` is loaded whenever you use `library(tidyverse)`,
but you can also load it individually:

```{r load_ggplot_again, options}
library(ggplot2)

cow = carData::Cowles
```

## Using `ggplot2` properly

`ggplot2` will always work best if:

* Everything you want to show in the plot is
  in a single dataframe.
* Each aspect of the plot is represented by a single
  column.
* Each column has the right data type, depending
  on whether it's categorical or continuous
  (`factor` or `numeric`).

You can make plots work if you don't have have those
things, but they will be much clunkier to put together. 
Take the time to put the data into the right shape
before you try to produce a plot.

## Mapping aesthetics

The most important concept in `ggplot2` is **mapping**
each variable in the data to an **aesthetic** feature
of the plot. A wide variety of plots can be put together
using a small number of common aesthetics. The aesthetic
mapping for a plot is set up using the `aes()` function.

Think about a scatterplot - one variable is mapped
to the position along the x-axis, and another
is mapped to the position along the y-axis. So
a basic scatterplot in ggplot is:

```{r basic_ggplot_scatterplot, fig.width=4, fig.height=3}
ggplot(cow, aes(x = neuroticism, y = extraversion)) +
    geom_point()
```

You could also map another variable to the colour of the points:

```{r ggplot_scatter_colours, fig.width=4, fig.height=3}
ggplot(cow, 
       aes(x = neuroticism, y = extraversion, colour = sex)) +
    geom_point()
```

Sometimes you want to change the look of *everything* in the plot,
without it being specific to individual data points. In that
case, you don't put it in the `aes()` mapping. Everything
that's connected to the data should still go in `aes()`. For
example, when scatterplots have lots of points it helps
to make them transparent so you can see how many points
are overlapping:

```{r ggplot_scatter_alpha, fig.width=4, fig.height=3}
ggplot(cow, 
       aes(x = neuroticism, y = extraversion, colour = sex)) +
    geom_point(alpha = 0.4)
```

## geoms: representing the data with different components

`ggplot2` doesn't have functions for specific kinds of plots
like a "boxplot" or "bar chart". It does have all the components
you need to put these together though.

A boxplot will usually have:

* A categorical variable mapped to `x`
* A continuous variable mapped to `y`
* Possibly extra variables mapped to the "fill colour" of
  the boxes.
  
```{r ggplot_boxplot_example, fig.width = 4, fig.height=3}
ggplot(cow, aes(x = sex, y = extraversion, fill = volunteer)) +
    geom_boxplot()
```

A bar chart will usually have:

* A discrete variable mapped to `x`
* Another variable mapped to `fill` that defines
  the subgroups.
* No variable mapped to `y`: `y` in a bar chart is
  normally just the count (number of rows) in
  each subgroup.
* The bars for the subgroups either "stacked" on top
  of each other or side by side (ggplot2 calls this
  "dodging")
  
```{r ggplot_bar_example, fig.width=4, fig.height=3}
ggplot(cow, aes(x = sex, fill = volunteer)) +
    geom_bar(position = "dodge")
```

We've done some very standard plot types here
to help illustrate how each plot can be broken down
into its aesthetic mappings, but the real power
of ggplot is that you aren't tied to the standard
kinds of plots - you can mix and match their
components to produce a unique plot that shows
exactly what you want.

## Combining geoms

Here we'll produce a plot that shows:

* The individual data points ("jittered" a bit so they
  don't all overlap)
* Contours that show how dense the points are
* A "line of best fit" that shows the overall relationship
  between two variables
  
When you plot multiple geoms, the order you put them
together defines which element will be on top - the
first geom is on the bottom.

```{r ggplot_complex_example, message=FALSE}
ggplot(cow, aes(x = neuroticism, y = extraversion)) +
    stat_density_2d(aes(fill = stat(level)),
                    geom = "polygon",
                    alpha = 0.6) +
    geom_jitter(alpha = 0.5) +
    geom_smooth(colour = "red") +
    scale_fill_viridis_c() +
    theme_bw()
```

This plot is probably a little bit too "busy" to be really useful -
it's up to you to figure out what aspects of the data are most useful
to represent in a plot.

## Other packages use `ggplot2` too

